Importance of depth refugia for reef resilience and intervention on the Great Barrier Reef under future climate change

### 2025-10-10 paper/paper_draft.qmd
Updated abstract

---
title: Importance of depth refugia for reef resilience and intervention on the Great Barrier Reef under future climate change
author:
  - name: Benjamin Grier
    email: grierben777@gmail.com
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Townsville, Queensland, Australia
    attributes:
        corresponding: true
  - name: Takuya Iwanaga
    email: t.iwanaga@aims.gov.au
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Townsville, Queensland, Australia
  - name: Pedro Ribeiro De Almeida
    email: p.ribeirodealmeida@aims.gov.au
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Townsville, Queensland, Australia
  - name: Chinenye Ani
    email: c.ani@aims.gov.au
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Townsville, Queensland, Australia
  - name: Sam Matthews
    email: s.matthews@aims.gov.au
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Townsville, Queensland, Australia
abstract: |
    Coral reefs globally face unprecedented threats from climate change, with the Great 
    Barrier Reef (GBR) experiencing cumulative stressors and increasingly severe declines 
    in coral cover from thermal stress events. Understanding drivers behind reef resilience 
    to climate impacts is critical for conservation planning and intervention strategies. A 
    scenario-based population modelling approach was adopted with larval connectivity dynamics, 
    thermal adaptation, and environmental factors to assess coral reef resilience across GBR 
    locations under future climate projections. Using five Global Climate Models (GCMs) 
    spanning five Equilibrium Climate Sensitivities,  reefs were analysed for coral cover 
    time series and their ability of reefs to maintain positive carbonate production budgets 
    under future conditions. Contrary to expectations, larval connectivity patterns did not 
    correlate with increases in maintenance of positive carbonate budgets. Instead, reef depth 
    emerged as the primary predictor of cover-time series resilience and maintenance of positive 
    carbonate budgets, with deeper reefs benefiting from thermal refugia that reduce bleaching 
    impacts. These findings suggest that depth provides critical 
    protection against thermal stress, contrasting current intervention practices that primarily target easily 
    accessible sites in the 1-10m depth range. These results have important implications 
    for reef management, indicating that depth refugia should be prioritized for further 
    consideration in conservation planning alongside connectivity modelling advancements 
    to maximize coral survival under continuing climate change. 
keywords:
  - Coral reef
  - Climate change
  - Ecosystem resilience
  - Reef depth
  - Reef connectivity
date: last-modified
format:
  elsevier-docx:
    journal:
      name: Journal Name
      formatting: preprint
      cite-style: authoryear
bibliography: references.bib
engine: julia
crossref:
  custom:
  - kind: float
    reference-prefix: SFigure
    key: sfig
  - kind: float
    reference-prefix: STable
    key: stbl
editor:
  markdown:
    wrap: 92
    canonical: true
---

>>>>

---
title: Importance of depth refugia for reef resilience and intervention on the Great Barrier Reef under future climate change
author:
  - name: Benjamin Grier
    email: grierben777@gmail.com
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Townsville, Queensland, Australia
    attributes:
        corresponding: true
  - name: Takuya Iwanaga
    email: t.iwanaga@aims.gov.au
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Townsville, Queensland, Australia
  - name: Pedro Ribeiro De Almeida
    email: p.ribeirodealmeida@aims.gov.au
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Townsville, Queensland, Australia
  - name: Chinenye Ani
    email: c.ani@aims.gov.au
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Townsville, Queensland, Australia
  - name: Sam Matthews
    email: s.matthews@aims.gov.au
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Townsville, Queensland, Australia
abstract: |
    Coral reefs globally face unprecedented threats from climate change, with the Great
    Barrier Reef (GBR) experiencing cumulative stressors and increasingly severe declines
    in coral cover from thermal stress events. Understanding drivers behind reef resilience
    to climate impacts is critical for conservation planning and intervention strategies. A
    scenario-based population modelling approach was adopted with larval connectivity dynamics and
    environmental factors to assess coral reef resilience across the GBR using projected conditions under
    five Global Climate Models (GCMs) spanning five Equilibrium Climate Sensitivities. Projected coral
    cover were analysed for each reefs ability to maintain positive carbonate production budgets under
    future conditions using coral cover as a proxy. Larval connectivity patterns did not correlate with
    increases in maintenance of positive carbonate budgets. Instead, reef depth emerged as the primary
    predictor, with deeper reefs benefitting from thermal refugia that reduce bleaching impacts. These
    findings suggest that depth is a tangible and pragmatic reef characteristic to consider in future
    intervention practices for coral reef restoration. These results have important implications for reef
    management, indicating that depth refugia should be prioritized for further consideration in
    conservation planning to maximize coral survival under continuing climate change.

keywords:
  - Coral reef
  - Climate change
  - Ecosystem resilience
  - Reef depth
  - Reef connectivity
date: last-modified
format:
  elsevier-docx:
    journal:
      name: Journal Name
      formatting: preprint
      cite-style: authoryear
bibliography: references.bib
engine: julia
crossref:
  custom:
  - kind: float
    reference-prefix: SFigure
    key: sfig
  - kind: float
    reference-prefix: STable
    key: stbl
editor:
  markdown:
    wrap: 92
    canonical: true
---
